### 2018-04-26 1_intro.tex
add minor fixes and multiple todo-comments.

@@

\end{tabular}
\caption{Store buffering (SB): a litmus test on write-read reordering allowed under the x86-TSO weak memory model}
\label{simple_wmm_x86}
\end{figure}

@@

Although weak memory studies is rather young research area, there exist frameworks and tools for exploring WMMs and examining simple programs with respect to the them. The state-of-the-art tool is \tool{diy} (for \textit{do it yourself}), developed by the researchers from INRIA institute, France and University of Cambridge, UK.
The \tool{diy}\footnote{Project web page: \url{diy.inria.fr}} is a software suite for designing and testing weak memory models. It is firstly released back in 2010, and since that time it remained to be the only tool for testing weak memory models. The \tool{diy} consists of several modules: the litmus tests generators \tool{diy}, \tool{diycross} and \tool{diyone}, the litmus tests concrete executor \tool{litmus} that runs tests on a physical machine while collecting its behaviours, and the weak memory models simulator \tool{herd} that implements reachability analysis for exploring states reachable under specified~WMM.

@@

The Chapter~\ref{ch:wmm} gives more detailed description of the weak memory model-aware analysis and provides description of memory models for some common architectures (x86, ARM and POWER, Sparc, ???). Chapter~...

@@

>>>>

%TODO: rename mousquetaires: http://www.cprover.org/wmm/musketeer/ !!! (+reference this project: "Don’t sit on the fence".pdf in alglave/ )
% TODO: see how musketeer deals with recursive calls: http://www.cprover.org/wmm/musketeer/rec.pdf

@@

\end{tabular}
\caption{Store buffering (SB): a litmus test on write-read reordering allowed under the x86-TSO and forbidden under the SC memory model}
\label{simple_wmm_x86}
\end{figure}

@@

Although weak memory studies is rather young research area, there exist frameworks and tools for exploring WMMs and examining simple programs with respect to the them. The state-of-the-art tool is \tool{diy} (for \textit{do it yourself}), developed by the researchers from INRIA institute, France and University of Cambridge, UK.
The \tool{diy}\footnote{Project web site: \url{http://diy.inria.fr/}} is a software suite for designing and testing weak memory models. It is firstly released back in 2010, and since that time it remained to be the only tool for testing weak memory models. The \tool{diy} consists of several modules: the litmus tests generators \tool{diy}, \tool{diycross} and \tool{diyone}, the litmus tests concrete executor \tool{litmus} that runs tests on a physical machine while collecting its behaviours, and the weak memory models simulator \tool{herd} that implements reachability analysis for exploring states reachable under specified~WMM.

@@

The thesis is organised as following. Chapter~\ref{ch:wmm} gives a general view on the weak memory model-aware analysis. Chapter~...

@@

% TODO: other tools for bounded model-checking: 
%Clarke's goto-cc : \cite{clarke2004tool} : http://www.cprover.org/goto-cc/
% (perhaps, goto-cc is a part of cprover) : cprover https://www.cprover.org/cbmc/doc/manual.pdf
%cbmc http://www.cprover.org/cbmc/

### 2018-04-26 2_wmm.tex
add minor fixes and multiple todo-comments.

In general, analysis of concurrent programs with respect to axiomatic memory models is performed in several stages. Firstly, the control-flow and data-flow of a program is encoded as the set of possible \textit{candidate executions}. %possible communications between threads of the program.
Obtained model of the program is called an anarchic semantics, which is a truly parallel semantics with no global time that describes all possible computations with all possible communications~\cite{alglave2016syntax}. Thereafter, the anarchic semantics is constrained by the \textit{weak memory model} specification which is a set of axiomatic constraints for filtering out executions inconsistent in particular architecture.

@@

@@

@@

\vspace{1em}
The basic relations are the following~\cite{alglave2010shared}:

\begin{itemize}
    \item The \textit{control-flow} of a program is defined by the \textit{program-order} relation \po~$\subset~\mathbb{E}~\times~\mathbb{E}$, which represents the total order of events of same process.
	For instance, if the instruction $i_1$ generates the event $e_1$ and the instruction $i_2$ follows $i_1$ and generates the event $e_2$, then $e_1 \relation{po} e_2$.
	%Some new relations may be acquired : dp, po-loc

@@

%\vspace{0.5em}
Below we enumerate some derived relations~\cite{alglave2010shared}:

@@

    \item the \textit{data dependency} relation \texttt{dp}, which is a subset of \po-relation that always has a read at its source (it connects the read to the write which it depends on).
    %$\mathtt{com} = (\mathtt{rf} \cup \mathtt{co} \cup \mathtt{fr})$

@@

The work~\cite{alglave2010shared} (Chapter~2) provides definition of basic properties of relations, such as \textit{reflexivity} and \textit{irreflexivity}, \textit{transitivity} and \textit{transitive closure}, \textit{acyclicity}.
Thereafter, weak memory models make asserts over these properties, thus restricting the set of allowed behaviours of the system.

@@

\begin{itemize}
	\item \textit{the union} of two relations \relname{r1} and \relname{r2} is \relname{r1\,|\,r2}
	\item \textit{the intersection} of two relations \relname{r1} and \relname{r2} is \relname{r1\,\&\,r2}
	\item \textit{the difference} of two relations \relname{r1} and \relname{r2} is $\mathtt{r1\,\backslash\,r2}$
	\item \textit{the sequence}\footnote{The sequence of two relations \relname{r1} and \relname{r2} is defined as the set of pairs \relname{(x,y)} such that there exists an intervening \relname{z}, such that $\relname{(x,z)} \in \relname{r1}$ and $\relname{(z,y)} \in \relname{r2}$}
	of two relations \relname{r1} and \relname{r2} is \relname{r1;r2}
\end{itemize}

For instance, Figure~\ref{example:x86-cat} contains part of x86-TSO model~\cite{owens2009better} that asserts acyclicity of communication relation and \relname{po-loc} relation:

@@

%// exmaple with reordering
%// ex. with
%// Rev-29 Example 7-6. Stores Are Transitively Visible. %see http://www.cl.cam.ac.uk/~pes20/weakmemory/x86tso-paper.pdf

%There is a barrier instruction \relation{mfence} that may be used for flushing the buffers into the main memory.

%briefly some hw memory models: X86-TSO, Alpha, POWER, ;

%language memory models: Java, C++;

%library-level kernel memory model, ref to github with tests

@@

%\section{Some examples of WMM}

%//axioms of TSO wmm

%// example of sets: rf, co, ... for the code snipped used before

>>>>

In general, analysis of concurrent programs with respect to axiomatic memory models is performed in several stages. %TODO: site the 'Weak Memory Models with Matching Axiomatic and Operational Definitions' https://arxiv.org/pdf/1710.04259.pdf
%TODO: cite also: "DENNIS SHASHA - Efficient and Correct Programs that Share" http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.142.2238&rep=rep1&type=pdf
Firstly, the control-flow and data-flow of a program is encoded as the set of possible \textit{candidate executions}. %possible communications between threads of the program.
Obtained model of the program is called an anarchic semantics, which is a truly parallel semantics with no global time that describes all possible computations with all possible communications~\cite{alglave2016syntax}. Thereafter, the anarchic semantics is constrained by the \textit{weak memory model} specification which is a set of axiomatic constraints for filtering out executions inconsistent in particular architecture.

@@

% TODO: the event-based representation is a non-deterministic model

% TODO: the semantics of the event-flow graph is given:  "alglave: Don’t sit on the fence A static analysis approach to automatic fence insertion" page 8 or 9
%"To build the aeg , we define a semantics of goto-programs in terms of abstract events, ..."
% TODO: OR from here: http://www.cprover.org/etaps/esop.pdf "Software Verification for Weak Memory via Program Transformation"

@@

% TODO: more fences: "On Power and ARM (see Sec. 6), we distinguish between control fence (which we write cfence), lightweight fence (lwfence) and full fence (ffence). On x86, implementing TSO, there is only one fence, called mfence." (herding cats, p. 14)

@@

\vspace{1em}
The basic relations are the following~\cite{alglave2010shared}:
%TODO: say that po and rf are  defined from the candidate execution:
%TODO: say that 
\begin{itemize}
    \item The \textit{control-flow} of a program is defined by the \textit{program-order} relation \po~$\subset~\mathbb{E}~\times~\mathbb{E}$, which represents the total order of events of same process.
	For instance, if the instruction $i_1$ generates the event $e_1$ and the instruction $i_2$ follows $i_1$ and generates the event $e_2$, then $e_1 \relation{po} e_2$.
	%Some new relations may be acquired : dp, po-loc

@@

%TODO: add ppo = characteristics of memory model. "Architectures are instances of our model. An architecture is a triple of functions (ppo, fences, prop), which specifies the preserved program order ppo, the fences fences and the propagation order prop." (herding cats, p.13)

Below we enumerate some derived relations~\cite{alglave2010shared}:

@@

    \item the \textit{data dependency} relation \texttt{dp}, which is a subset of \po-relation that always has a read at its source (it connects the read to the write which it depends on).
    %TODO: say that po-relation is *static*
    %$\mathtt{com} = (\mathtt{rf} \cup \mathtt{co} \cup \mathtt{fr})$

@@

The work~\cite[Chapter~2]{alglave2010shared} provides definition of basic properties of relations, such as \textit{reflexivity} and \textit{irreflexivity}, \textit{transitivity} and \textit{transitive closure}, \textit{acyclicity}.
Thereafter, weak memory models make asserts over these properties, thus restricting the set of allowed behaviours of the system.

@@

\begin{itemize}
	\item \textit{the union} of two relations \rel{r1} and \rel{r2} is \rel{r1\,|\,r2}
	\item \textit{the intersection} of two relations \rel{r1} and \rel{r2} is \rel{r1\,\&\,r2}
	\item \textit{the difference} of two relations \rel{r1} and \rel{r2} is $\mathtt{r1\,\backslash\,r2}$
	\item \textit{the sequence}\footnote{The sequence of two relations \rel{r1} and \rel{r2} is defined as the set of pairs \rel{(x,y)} such that there exists an intervening \rel{z}, such that $\rel{(x,z)} \in \rel{r1}$ and $\rel{(z,y)} \in \rel{r2}$}
	of two relations \rel{r1} and \rel{r2} is \rel{r1;r2}
\end{itemize}

For instance, the \fr-relation is defined as follows: $\mathr{fr} = (\mathr{rf}^{-1};\mathr{co})$.

Figure~\ref{example:x86-cat} contains part of x86-TSO model~\cite{owens2009better} that asserts acyclicity of communication relation and \rel{po-loc} relation:

@@

@@

### 2018-04-26 3_encod.tex
Consider Prof. Heljanko's comments.
add minor fixes and multiple todo-comments.

%\chapter{SMT-based program analysis}
%\chapter{Portability of concurrent software}
\chapter{Portability analysis as a SAT~problem}
\label{ch:port}

@@

\section{Model checking}
\label{ch:port:mc}

The classical model checking algorithms explore the state space of an abstract automata or a transition system in order to find states that violate the specification. The general schema of model checking is the following: firstly, the analysing system is being represented as a transition system, a finite directed graph with labeled nodes representing states of the system such that each state corresponds to the unique subset of atomic propositions, that characterise the behavioral properties of each state. 
Then, the system constraints are being defined in terms of a modal temporal logic with respect to the atomic propositions. Commonly, the Linear Temporal Logic~(LTL) or Computational Tree Logic~(CTL), along with their extensions, are used as a specification language due to the expressiveness and verifiability of their statements. 
In the described schema, the model checking problem is reducible to the reachability analysis, an iterative process of a systematic exhaustive search in the state space. This approach is called \textit{unbounded model checking (UMC)}.

However, all model checking techniques are exposed to the \textit{state explosion problem} as the size of the state space grows exponentially with respect to the number of state variables of the system. In case of modeling concurrent systems, this problem becomes much more considerable due to exponential number of possible interleavings of states.
Therefore, the research in model checking over past 40 years was aimed at tackling the state explosion problem, mostly by optimising search space, search strategy or basic data structures of existing algorithms.

@@

As we have said before, a program $P$ is called portable from the source architecture (defined as a memory model) $\mathcal{M_S}$ to the target architecture $\mathcal{M_T}$ if \textit{all} executions consistent under $\mathcal{M_T}$ are consistent under $\mathcal{M_S}$~\cite{Porthos17}:

@@

Note that the definition of portability requirements against \textit{executions} is strong enough, as it implies the portability against \textit{states} (the \textit{state-portability})~\cite{Porthos17}.

%It is possible to formulate this requirement as an SMT formula, so that the portability analysis problem becomes reduced to the bounded reachability problem.
The result SMT formula $\phi$ of the portability problem should contain both encodings of control-flow $\phi_{CF}$ and data-flow $\phi_{DF}$ of the program, and assertions of both memory models: $\phi = \phi_{CF} \land \phi_{DF} \land \phi_{\mathcal{M_T}} \land \phi_{\lnot\mathcal{M_S}}$. If the formula is satisfiable, there exist a portability bug.
%The control-flow and data-flow encodings are standard for BMC~\cite{collavizza2006exploration}, they are described below. 
%However, encoding of memory models requires additional techniques due to recursive definitions of relations, that were proposed in~\cite{Porthos17}.

@@

Let $\fx : \mathbb{E} \rightarrow \{0,1\}$ be the predicate that signifies the fact that the event has been e\textbf{x}ecuted (and, consequently, has changed the state of the system).
Let $\fv : \mathbb{C} \rightarrow \mathbb{N}$ be the function that returns the value of the computation event that will be computed once the event is executed (strictly speaking, it retuns the \textit{set} of values determined by the \relation{rf}-relation; see Chapter~? for the relations encoding%TODO
). We distinguish the function $\fv_p : \mathbb{C_p} \rightarrow \{0,1\}$ that evaluates the predicative computation event. In the result formula, all symbols $\fx(e_i)$ and $\fv(e_i)$ are encoded as boolean variables.

@@

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.3\textwidth}
        \makebox[\textwidth]{
        \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.5cm,semithick]
            \node[c] (1) [] {$e_1$};
            \node[c] (2) [below of=1] {$e_2$};
            \path[->]
            (1) edge [] node {} (2)
            ;
        \end{tikzpicture}
        }
        \caption{Sequential events}
        \label{encode:cf:seq}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.35\textwidth}
        \makebox[\textwidth]{
        \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.5cm,semithick]
            \node[c] (1) [] {$e_1$};
            \node[c] (2) [below left of=1] {$e_2$};
            \node[c] (3) [below right of=1] {$e_3$};
            \path[->]
            (1) edge [] node {} (2)
            (1) edge [dotted] node {} (3)
            ;
        \end{tikzpicture}
        }
        \caption{Conditional branching}
        \label{encode:cf:br}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.25\textwidth}
        \makebox[\textwidth]{
        \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.5cm,semithick]
            \node[c] (i) [] {$e_i$};
            \node[c,draw=none] (ii) [right=0.2cm of i] {};
            \node[c,draw=none] (iii) [right=0.2cm of ii] {$...$};
            \node[c,draw=none] (iv) [right=0.2cm of iii] {};
            \node[c] (v) [right=0.2cm of iv] {$e_{i+j}$};
            \node[c] (k) [below of=iii] {$e_k$};
            \path[->]
            (i) edge [] node {} (k)
            (ii) edge [] node {} (k)
            (iii) edge [dotted] node {} (k)
            (iv) edge [dotted] node {} (k)
            (v) edge [] node {} (k)
            ;
        \end{tikzpicture}
        }
        \caption{Branch merging}
        \label{encode:cf:merge}
    \end{subfigure}
    %
    \caption{Linear and non-linear cases of control-flow graph}
    \label{encode:cf}
\end{figure}

For listed cases, below we propose the encoding scheme that uniquely encodes each node of graph and allows to encode partially executed program.
The Equation~\ref{enc:cf:seq} for encoding the sequential control-flow represented in Figure~\ref{encode:cf:seq} reflects the fact that the event $e_2$ can be executed iff the event $e_1$ was executed. The Equation~\ref{enc:cf:br} for encoding the branching control-flow depicted in Figure~\ref{encode:cf:br} allows only following executions: $\{\emptyset, (e_1), (e_1 \rightarrow e_2), (e_1 \rightarrow e3) \}$. In encoding~\ref{enc:cf:merge} of the merge-point represented in Figure~\ref{encode:cf:merge}, the event $e_k$ is executed if either of its predecessors was executed, regardless of type of the transition.
%TODO: maybe add the overall formula \phi_{CF} ?

\begin{align}
\phi_{CF_{seq}}   = \ & \fx(e_2) \rightarrow \fx(e_1) \label{enc:cf:seq} \\
\phi_{CF_{br}}    = \ & [\fx(e_2) \rightarrow \fx(e_1)] \land [\fx(e_3) \rightarrow \fx(e_1)] \land \nonumber \\
				  &  [\fx(e_2) \rightarrow \fv(e_1)] \land [\fx(e_3) \rightarrow \lnot\fv(e_1)] \land \nonumber \\
				  & \lnot [\fx(e_2) \land \fx(e_3)]  \label{enc:cf:br} \\
\phi_{CF_{merge}} = \ & \fx(e_k) \rightarrow (\bigvee\limits_{e_p \in \mathtt{pred}(e_k)}^{} \fx(e_p)) \label{enc:cf:merge}
\end{align}

For the sake of encoding correctness, we require all branches to have at least one event. Thus, for branching statements that do not have any events in one of the branches (such a branch represents a conditional jump forward), we add the synthetic \texttt{nop}-event as it is shown in Figure~\ref{encode:branching:nop}:

@@

Following~\cite{Porthos17}, the indexed references of variables are computed in accordance with the following rules:
(1)~any access to a shared variable (both read and write) increments its SSA-index;
(2)~only writes to a local variable increment its SSA-index (reads preserve indices);
(3)~no access to a constant variable or computed expression changes their SSA-index.
These rules determine the following encoding of load, store and computation events within single thread:
%TODO: maybe add the overall formula \phi_{DF} ?
\begin{align}
    \phi_{DF_{e = \mathtt{load}(r \leftarrow l)}} = \ & \fx(e) \rightarrow (r_{i+1} = l_{i+1}) \\
    \phi_{DF_{e = \mathtt{store}(l \leftarrow r)}} = \ & \fx(e) \rightarrow (l_{i+1} = r_i) \\
    \phi_{DF_{e = \mathtt{compute}(...)}} = \ & \fx(e) \rightarrow \fv(e) % \\
\end{align}

To convert the program into SSA form, for each event each variable that is declared so far (either local or shared) is mapped to its indexed reference; this information is stored in the SSA-map "event \textit{to} variable \textit{to} SSA-index". %for each event stores the map that each declared variable maps to its SSA-index
%"$\{$ event : $\{$ variable : index $\}\}$",
%"$\left\{$ event : $\left\{$ variable : index $\right\} \right\}$",
The SSA-map is computed iteratively while traversing the event-flow graph in topological order as it is described in Algorithm~\ref{algorithm:ssa-map}:

\begin{algorithm}
    \caption{Algorithm for computing the SSA-indices}\label{alg:compute-ssa}
    \algorithmicrequire{The event-flow graph $G = \langle N, E \rangle$ where $V$ is the set of nodes (events), $E$ is the set of control-flow transitions, $e_0$ is the entry node}
    \algorithmicensure{The SSA-map of the form "$\{$ event : $\{$ variable : index $\}\}$"}
    \begin{algorithmic}[1]
        \Function{ComputeSSA}{G}
            \State {$S \gets$ empty map; $S[e_0] \gets$ empty map}
            \ForAll {event $e_i \in G.N$ in topological order}
                \ForAll {predecessor $e_j \in \mathtt{pred}(e_i)$}
                    \State {$S[e_i] \gets \mathtt{copy}(S[e_j])$}
                    \ForAll {variable $v_k \in$ set of variables accessed by $e_i$ }
                        \State {$S[e_i][v_k] \gets \mathtt{max}(S[e_i][v_k], \ S[e_j][v_k])$}
                        \If {need to update the index of $v_k$} \Comment {cases (1)-(2)}
                            \State {$S[e_i][v_k] \gets S[e_i][v_k] + 1$}
                        \EndIf
                    \EndFor
                \EndFor
            \EndFor
        \EndFunction
    \end{algorithmic}
    \label{algorithm:ssa-map}
\end{algorithm}

@@

%The encoding of a weak memory model itself does not depend on the analysing program. The problem here...

recursive definitions and Kleene iteration

As it is described in~\cite{Porthos17}, the is based . . .
%~\cite{Porthos17}%, Section "Programs and Memory Models", subsection "Encoding Derived Relations"

>>>>

%\chapter{SMT-based program analysis}
%\chapter{Portability of concurrent software}
\chapter{Portability analysis as an SMT~problem}
\label{ch:port}

@@

\section{Model checking and reachability analysis}
\label{ch:port:mc}

The model checking is the problem of verifying the system (the model) against the set of constraints (the specification).
As the state machine model is the most widespread mathematical model of computation, most classical model checking algorithms explore the state space of a system in order to find states that violate the specification.
The general schema of model checking is the following: firstly, the analysing system is being represented as a transition system, a finite directed graph with labeled nodes representing states of the system such that each state corresponds to the unique subset of atomic propositions, that characterise the behavioral properties of each state.
Then, the system constraints are being defined in terms of a modal temporal logic with respect to the atomic propositions. Commonly, the Linear Temporal Logic~(LTL) or Computational Tree Logic~(CTL), along with their extensions, are used as a specification language due to the expressiveness and verifiability of their statements. 
In the described schema, the model checking problem is reducible to the reachability analysis, an iterative process of a systematic exhaustive search in the state space. This approach is called \textit{unbounded model checking (UMC)}.

However, all model checking techniques are exposed to the \textit{state explosion problem} as the size of the state space grows exponentially with respect to the number of state variables used by the system (its size). In case of modeling concurrent systems, this problem becomes much more considerable due to exponential number of possible interleavings of states.
Therefore, the research in model checking over past 40 years was aimed at tackling the state explosion problem, mostly by optimising search space, search strategy or basic data structures of existing algorithms.

@@

Consider the function $cons_{\mathcal{M}}(P)$ calculates the set of executions of program $P$ consistent under the memory model $\mathcal{M}$. Then, the program $P$ is called portable from the source architecture (memory model) $\mathcal{M_S}$ to the target architecture $\mathcal{M_T}$ if all executions consistent under $\mathcal{M_T}$ are consistent under $\mathcal{M_S}$~\cite{Porthos17}:

@@

Note that the definition of portability requirements against \textit{executions} is strong enough, as it implies the portability against \textit{states} (the \textit{state-portability})~\cite{Porthos17}.
The result SMT formula $\phi$ of the portability problem should contain both encodings of control-flow $\phi_{CF}$ and data-flow $\phi_{DF}$ of the program, and assertions of both memory models: $\phi = \phi_{CF} \land \phi_{DF} \land \phi_{\mathcal{M_T}} \land \phi_{\lnot\mathcal{M_S}}$. If the formula is satisfiable, there exist a portability bug.
%The control-flow and data-flow encodings are standard for BMC~\cite{collavizza2006exploration}, they are described below. 
%However, encoding of memory models requires additional techniques due to recursive definitions of relations, that were proposed in~\cite{Porthos17}.

@@

Let $\fx : \mathbb{E} \rightarrow \{0,1\}$ be the predicate that signifies the fact that the event has been e\textbf{x}ecuted (and, consequently, has changed the state of the system).
Let $\fv : \mathbb{C} \rightarrow \mathbb{R}$ be the function that returns the value of the computation event (evaluates it) that will be computed once the event is executed (strictly speaking, it retuns the \textit{set} of values determined by the \relation{rf}-relation; see Chapter~\textbf{?TODO?} for the relations encoding%TODO
). We distinguish the function $\fv_p : \mathbb{C_p} \rightarrow \{0,1\}$ that evaluates the predicative computation event. In the result formula, all symbols $\fx(e_i)$ and $\fv(e_i)$ are encoded as boolean variables.

@@

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.3\textwidth}
        \makebox[\textwidth]{
        \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.5cm,semithick]
            \node[c] (1) [] {$e_1$};
            \node[c] (2) [below of=1] {$e_2$};
            \path[->]
            (1) edge [] node {} (2)
            ;
        \end{tikzpicture}
        }
        \caption{Sequential events (\texttt{seq})}
        \label{encode:cf:seq}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.35\textwidth}
        \makebox[\textwidth]{
        \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.5cm,semithick]
            \node[c] (1) [] {$e_1$};
            \node[c] (2) [below left of=1] {$e_2$};
            \node[c] (3) [below right of=1] {$e_3$};
            \path[->]
            (1) edge [] node {} (2)
            (1) edge [dotted] node {} (3)
            ;
        \end{tikzpicture}
        }
        \caption{Conditional branching (\texttt{br})}
        \label{encode:cf:br}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.25\textwidth}
        \makebox[\textwidth]{
        \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.5cm,semithick]
            \node[c] (i) [] {$e_i$};
            \node[c,draw=none] (ii) [right=0.2cm of i] {};
            \node[c,draw=none] (iii) [right=0.2cm of ii] {$...$};
            \node[c,draw=none] (iv) [right=0.2cm of iii] {};
            \node[c] (v) [right=0.2cm of iv] {$e_{i+j}$};
            \node[c] (k) [below of=iii] {$e_k$};
            \path[->]
            (i) edge [] node {} (k)
            (ii) edge [] node {} (k)
            (iii) edge [dotted] node {} (k)
            (iv) edge [dotted] node {} (k)
            (v) edge [] node {} (k)
            ;
        \end{tikzpicture}
        }
        \caption{Branch merging (\texttt{mer})}
        \label{encode:cf:merge}
    \end{subfigure}
    %
    \caption{Linear and non-linear cases of control-flow graph}
    \label{encode:cf}
\end{figure}

For listed cases, below we propose the encoding scheme that uniquely encodes each node of graph and allows to encode partially executed program.
Equation~\ref{enc:cf:seq} encodes the sequential control-flow represented in Figure~\ref{encode:cf:seq} and reflects the fact that the event $e_2$ can be executed iff the event $e_1$ has been executed.
Equation~\ref{enc:cf:br} encodes the branching control-flow depicted in Figure~\ref{encode:cf:br} by allowing only following executions: $\{\emptyset, (e_1), (e_1~\rightarrow~e_2), (e_1 ~\rightarrow~e_3)\}$. In encoding~\ref{enc:cf:merge} of the merge-point represented in Figure~\ref{encode:cf:merge}, the event $e_k$ is executed if either of its predecessors was executed, regardless of type of the transition.
%TODO: maybe add the overall formula \phi_{CF} ?

\begin{align}
\phi_{CF_{seq}}   = \ & \fx(e_2) \rightarrow \fx(e_1) \label{enc:cf:seq} \\
\phi_{CF_{br}}    = \ & [\fx(e_2) \rightarrow \fx(e_1)] \land [\fx(e_3) \rightarrow \fx(e_1)] \land \nonumber \\
				  &  [\fx(e_2) \rightarrow \fv(e_1)] \land [\fx(e_3) \rightarrow \lnot\fv(e_1)] \land \nonumber \\
				  & \lnot [\fx(e_2) \land \fx(e_3)]  \label{enc:cf:br} \\
\phi_{CF_{mer}} = \ & \fx(e_k) \rightarrow (\bigvee\limits_{e_p \in \mathtt{pred}(e_k)}^{} \fx(e_p)) \label{enc:cf:merge}
\end{align}

For sake of encoding correctness, we require all branches to have at least one event.
Thus, for branching statements that do not have any events in one of the branches (such a branch represents a conditional jump forward), we add the synthetic \texttt{nop}-event as it is shown in Figure~\ref{encode:branching:nop}:

@@

Following~\cite{Porthos17}, the indexed references of variables are computed in accordance with the following rules:
(1)~any access to a shared variable (both read and write) increments its SSA-index;
(2)~only writes to a local variable increment its SSA-index (reads preserve indices);
(3)~no access to a constant variable or computed (evaluated) expression changes their SSA-index.
These rules determine the following encoding of load, store and computation events within single thread:
%TODO: maybe add the overall formula \phi_{DF} ?
\begin{align}
    \phi_{DF_{e = \mathtt{load}(r \leftarrow l)}} = \ & \fx(e) \rightarrow (r_{i+1} = l_{i+1}) \\
    \phi_{DF_{e = \mathtt{store}(l \leftarrow r)}} = \ & \fx(e) \rightarrow (l_{i+1} = r_i) \\
    \phi_{DF_{e = \mathtt{eval}(...)}} = \ & \fx(e) \rightarrow \fv(e) % \\
\end{align}

To convert the program into SSA form, for each event each variable that is declared so far (either local or shared) is mapped to its indexed reference; this information is stored in the SSA-map "event \textit{to} variable \textit{to} SSA-index". %for each event stores the map that each declared variable maps to its SSA-index
%"$\{$ event : $\{$ variable : index $\}\}$",
%"$\left\{$ event : $\left\{$ variable : index $\right\} \right\}$",
The SSA-map is computed iteratively while traversing the event-flow graph in topological order as it is described in Algorithm~\ref{algorithm:ssa-map}.

\begin{algorithm}
    \caption{Algorithm for computing the SSA-indices}\label{alg:compute-ssa}
    \algorithmicrequire{The event-flow graph $G = \langle N, E \rangle$ where $V$ is the set of nodes (events), $E$ is the set of control-flow transitions, $e_0$ is the entry node}
    \algorithmicensure{The SSA-map of the form "$\{$ event : $\{$ variable : index $\}\}$"}
    \begin{algorithmic}[1]
        \Function{Compute-SSA-map}{G}
            \State {$S \gets$ empty map; $S[e_0] \gets$ empty map}
            \ForAll {event $e_i \in G.N$ in topological order}
                \ForAll {predecessor $e_j \in \mathtt{pred}(e_i)$}
                    \State {$S[e_i] \gets \mathtt{copy}(S[e_j])$}
                    \ForAll {variable $v_k \in$ set of variables accessed by $e_i$ }
                        \State {$S[e_i][v_k] \gets \mathtt{max}(S[e_i][v_k], \ S[e_j][v_k])$}
                        \If {need to update the index of $v_k$} \Comment {cases (1)-(2)}
                            \State {$S[e_i][v_k] \gets S[e_i][v_k] + 1$}
                        \EndIf
                    \EndFor
                \EndFor
            \EndFor
        \EndFunction
    \end{algorithmic}
    \label{algorithm:ssa-map}
\end{algorithm}

@@

todo
%Most bounded model-checking problems are solved via computing the fixpoint for the

%The encoding of a weak memory model itself does not depend on the analysing program. The problem here...
%recursive definitions and Kleene iteration

%r 1 ∪r 2 (e 1 , e 2 ) ⇔ r 1 (e 1 , e 2 ) ∨ r 2 (e 1 , e 2 )
%r 1 ∩r 2 (e 1 , e 2 ) ⇔ r 1 (e 1 , e 2 ) ∧ r 2 (e 1 , e 2 )
%r 1 \r 2 (e 1 , e 2 ) ⇔ r W
%r −1 (e 1 , e 2 ) ⇔ r(e 2 , e 1 )
%1 (e 1 , e 2 ) ∧ ¬r 2 (e 1 , e 2 )
%r 1 (e 1 , e 3 ) ∧ r 2 (e 3 , e 2 )
%r ∗ (e 1 , e 2 ) ⇔ r + (e 1 , e 2 ) ∨ (e 1 = e 2 )
%r 1 ;r 2 (e 1 , e 2 ) ⇔
%e 3 ∈E
%r + (e 1 , e 2 ) ⇔ tc ⌈log |E|⌉ (e 1 , e 2 ), where
%tc 0 (e 1 , e 2 ) ⇔ r(e 1 , e 2 ), and
%tc i+1 (e 1 , e 2 ) ⇔ r(e 1 , e 2 ) ∨ (tc i (e 1 , e 2 ); tc i (e 1 , e 2 ))
%
%As it is described in~\cite{Porthos17}, the is based . . .
%~\cite{Porthos17}%, Section "Programs and Memory Models", subsection "Encoding Derived Relations"

### 2018-04-26 4_input.tex
Change the illustration for the litmus-test example.
add minor fixes and multiple todo-comments.

The \Porthos tool used a small subset of C as an input language as described on Figure~\ref{syntax:in_grammar_pts}~\cite{Porthos17}:

\begin{figure}[H]
%\begin{multicols}{2}
\begin{lstlisting}[mathescape=true,
        caption={Syntax of an input language of \Porthos},
        label={syntax:in_grammar_pts},
        literate={<--}{{$\leftarrow$}}1,
        morekeywords={if,then,else,return,while,program,thread,mfence,sync,lwsync,isync}
        breaklines=true,
        ]
<prog> : program <thrd>*
       ;
<thrd> : thread <tid> <inst>
       ;
<inst> : <atom> 
       | <inst> ; <inst>
       | while <pred> <inst>
       | if <pred > then <inst> else <inst>
       ;
<atom> : <reg> <-- <exp> 
       | <reg> <-- <loc>
       | <loc> := <reg> 
       | mfence
       | sync
       | lwsync
       | isync
       ;
\end{lstlisting}
%\end{multicols}
\end{figure}

The input language of \Porthos supported recursive definitions of control-flow instructions, that were directly encoded into SMT-formula. In \mousquetaires, the set of supported instructions is extended in order to support expressions valid in C.%TODO: examples, which  we extended
 Also, for the purpose of simplicity, all kinds of fence instructions and different types of variable assignments (shared variable assignment~`\texttt{:=}' or local variable assignment~`$\mathtt{\leftarrow}$') were defined directly in the grammar for \Porthos, whereas the semantics of function invocations need to be determined dynamically during the interpretation
% according to the set of known functions
(see Chapter~\ref{???}).

>>>>

%TODO: say sth about 100500 litmus-tests for kernel

The \Porthos tool used a small subset of C as an input language as described in Figure~\ref{syntax:in_grammar_pts}~\cite{Porthos17}.
It supports recursive definitions of control-flow instructions, that were directly encoded into SMT-formula.
In \mousquetaires, the set of supported instructions is extended in order to support expressions valid in C. %TODO: examples, how we extended
Also, for the purpose of simplicity, all kinds of fence instructions and different types of variable assignments (shared variable assignment~`\texttt{:=}' or local variable assignment~`$\mathtt{\leftarrow}$') were defined directly in the grammar for \Porthos, whereas the semantics of function invocations need to be determined dynamically during the interpretation
% according to the set of known functions
(see Chapter~\ref{???}).

\begin{figure}%[H]
%\begin{multicols}{2}
\begin{lstlisting}[mathescape=true,
        caption={Syntax of an input language of \Porthos},
        label={syntax:in_grammar_pts},
        literate={<--}{{$\leftarrow$}}1,
        morekeywords={if,then,else,return,while,program,thread,mfence,sync,lwsync,isync}
        breaklines=true,
        basicstyle=\ttfamily\scriptsize,
        ]
<prog> : program <thrd>*
       ;
<thrd> : thread <tid> <inst>
       ;
<inst> : <atom> 
       | <inst> ; <inst>
       | while <pred> <inst>
       | if <pred > then <inst> else <inst>
       ;
<atom> : <reg> <-- <exp> 
       | <reg> <-- <loc>
       | <loc> := <reg> 
       | mfence
       | sync
       | lwsync
       | isync
       ;
\end{lstlisting}
%\end{multicols}
\end{figure}
